Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2023-12-16 main.tex
:pencil: Add MIMO PC SV analysis

		Fig.~\ref{sm:pc_power_bond_hd} and \ref{sm:pc_power_bond_nd} compare the average channel power without and with direct link.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% ``Cascaded'' means the maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		``Cascaded'' means the \emph{power product} of the forward and backward channels.
		We observe that diagonal \gls{ris} wastes substantial cascaded power and struggles to align the direct-indirect eigenspace.
		When the direct link is absent, only \qty{2.6}{\percent} of available power is utilized by diagonal \gls{ris} while \qty{100}{\percent} power is recycled by fully-connected \gls{ris}.
		When the direct link is present, the proposed \gls{bd} \gls{ris} design can balance the direct-indirect and forward-backward eigenspace alignment for an optimal channel boost.
		It is worth noting that, when $L$ is sufficiently large, the composite channel power surpasses the power sum of direct and cascaded channels, thanks to the constructive \emph{amplitude superposition} of direct and cascaded channels.
		This again emphasizes the advantage of in-group connection of \gls{bd} \gls{ris}.
		% This is because the direct and indirect channels superpose
		% the proposed \gls{bd} \gls{ris} design can also be applied to conventional \gls{ris} to improve the channel power.
		% When the direct link is present, a small $L$ cannot align the direct-indirect eigenspace.

@@

	\begin{subsection}{Channel Singular Value Redistribution}
		We first show the channel shaping benefit of \gls{bd} \gls{ris} by a toy example.
		Consider $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (2, 2, 2)$ and assume the direct link is absent.
		The diagonal \gls{ris} is $\mathbf{\Theta}^\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the unitary \gls{ris} has 4 independent angular parameters
		\begin{equation}
			\mathbf{\Theta}^\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
		\end{equation}
		When the direct link is absent, $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		For a fair comparison, we enforce \emph{symmetry} with $\beta = \pi / 2$.
		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/pc_singular_toy.eps}
			\caption{Channel singular value shaping by diagonal and symmetry unitary \gls{ris} for $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (2, 2, 2)$ without direct link.}
			\label{sm:pc_singular_toy}
		\end{figure}
		Fig.~\ref{sm:pc_singular_toy} illustrates all possible channel singular values achieved by diagonal and symmetry unitary \gls{ris}.
		Despite using the same number of elements and parameters, \gls{bd} \gls{ris} provides much wider dynamic ranges of $\sigma_1(\mathbf{H})$ and $\sigma_2(\mathbf{H})$ than diagonal \gls{ris}.
		Larger gaps are expected when the symmetry constraint can be relaxed.

@@

		\begin{subsubsection}{Rank-1 Indirect Channel}
			The indirect channel is rank-1 iff the forward or backward channel is rank-1.
			Let $\mathbf{H}^\mathrm{F} = \sigma^\mathrm{F} \mathbf{u}^\mathrm{F} {\mathbf{v}^\mathrm{F}}^\mathsf{H}$ without loss of generality.
			In this case, the channel Gram matrix can be written as Hermitian-plus-rank-1:
			\begin{equation}
				\mathbf{G} \triangleq \mathbf{H} \mathbf{H}^\mathsf{H} = \mathbf{Y} + \mathbf{z} \mathbf{z}^\mathsf{H},
			\end{equation}
			where $\mathbf{Y} \triangleq \mathbf{H}^\mathrm{D} (\mathbf{I} - \mathbf{v}^\mathrm{F} {\mathbf{v}^\mathrm{F}}^\mathsf{H}) \mathbf{H}^\mathrm{D} = \mathbf{F} \mathbf{F}^\mathsf{H}$ and $\mathbf{z} \triangleq \sigma^\mathrm{F} \mathbf{H}^\mathrm{B} \mathbf{\Theta} \mathbf{u}^\mathrm{F} + \mathbf{H}^\mathrm{D} \mathbf{v}^\mathrm{F}$.
			Regardless of \gls{ris} size and structure\footnote{Similar conclusion was made for diagonal \gls{ris} in \cite{Semmler2023}.}, its $n$-th ($n \ge 2$) eigenvalues are bounded by the Cauchy interlacing formula \cite{Golub2013}
			\begin{equation}
				\lambda_1(\mathbf{Y}) \ge {\color{blue}\lambda_2(\mathbf{G})} \ge \lambda_2(\mathbf{Y}) \ge \ldots \ge \lambda_{n-1}(\mathbf{Y}) \ge {\color{blue}\lambda_n(\mathbf{G})} \ge \lambda_n(\mathbf{Y}).
			\end{equation}
			That is to say, [for sv]

			% \begin{equation}
			% 	\mathbf{G} = \underbrace{\mathbf{H}_\text{D} (\mathbf{I} - \mathbf{v}_{\text{F}} \mathbf{v}_{\text{F}}^\mathsf{H}) \mathbf{H}_\text{D}^\mathsf{H}}_{\mathbf{C}} +
			% 	\underbrace{
			% 		\begin{bsmallmatrix}
			% 			\sigma_{\text{F}} \mathbf{H}_\text{B} & \mathbf{H}_\text{D} \mathbf{v}_{\text{F}}
			% 		\end{bsmallmatrix}
			% 		\begin{bsmallmatrix}
			% 			\mathbf{\Theta} \mathbf{u}_{\text{F}} \\
			% 			1
			% 		\end{bsmallmatrix}
			% 	}_{\mathbf{z} = \sigma_{\text{F}} \mathbf{H}_\text{B} \mathbf{\Theta} \mathbf{u}_{\text{F}} + \mathbf{H}_\text{D} \mathbf{v}_{\text{F}}}
			% 	\begin{bsmallmatrix}
			% 		\mathbf{\Theta} \mathbf{u}_{\text{F}} \\
			% 		1
			% 	\end{bsmallmatrix}^\mathsf{H}
			% 	\begin{bsmallmatrix}
			% 		\sigma_{\text{F}} \mathbf{H}_\text{B} & \mathbf{H}_\text{D} \mathbf{v}_{\text{F}}
			% 	\end{bsmallmatrix}^\mathsf{H}
			% \end{equation}
		\end{subsubsection}

>>>>

		Fig.~\ref{sm:pc_power_bond_hd} and \ref{sm:pc_power_bond_nd} compare the average channel power without and with direct link.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% where the cascaded channel power available to passive \gls{ris} is $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% ``Cascaded'' means the maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% maximum power of the cascaded channel, i.e., $\lVert \mathbf{H}^\mathrm{B} \rVert _\mathrm{F}^2 \lVert \mathbf{H}^\mathrm{F} \rVert _\mathrm{F}^2$.
		% ? ``Cascaded'' means the \emph{power product} of the forward and backward channels.
		% ``Cascaded'' means the sum of element-wise product of first $N = \min(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R})$ eigenvalues of the forward and backward channels.
		``Cascaded'' means the sum of element-wise product of first $N = \min(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R})$ eigenvalues (i.e., element-wise power product) of the forward and backward channels.
		We observe that diagonal \gls{ris} wastes substantial cascaded power and struggles to align the direct-indirect eigenspace.
		When the direct link is absent, only \qty{2.6}{\percent} of available power is utilized by diagonal \gls{ris} while \qty{100}{\percent} power is recycled by fully-connected \gls{ris}.
		When the direct link is present, the proposed \gls{bd} \gls{ris} design can balance the direct-indirect and forward-backward eigenspace alignment for an optimal channel boost.
		It is worth noting that, when $L$ is sufficiently large, the composite channel power surpasses the power sum of direct and cascaded channels, thanks to the constructive \emph{amplitude superposition} of direct and cascaded channels.
		This again emphasizes the advantage of in-group connection of \gls{bd} \gls{ris}.
		% This is because the direct and indirect channels superpose
		% the proposed \gls{bd} \gls{ris} design can also be applied to conventional \gls{ris} to improve the channel power.
		% When the direct link is present, a small $L$ cannot align the direct-indirect eigenspace.

@@

	\begin{subsection}{Channel Singular Value Redistribution}
		We first show the channel shaping benefit of \gls{bd} \gls{ris} by a toy example.
		Consider $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (2, 2, 2)$ and assume the direct link is absent.
		The diagonal \gls{ris} is $\mathbf{\Theta}^\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the unitary \gls{ris} has 4 independent angular parameters
		\begin{equation}
			\mathbf{\Theta}^\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
		\end{equation}
		When the direct link is absent, $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		For a fair comparison, we enforce symmetry with $\beta = \pi / 2$.
		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/pc_singular_toy.eps}
			\caption{Channel singular value shaping by diagonal and symmetry unitary \gls{ris} for $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (2, 2, 2)$ without direct link.}
			\label{sm:pc_singular_toy}
		\end{figure}
		Fig.~\ref{sm:pc_singular_toy} illustrates all possible channel singular values achieved by diagonal and symmetry unitary \gls{ris}.
		Despite using the same number of elements and parameters, \gls{bd} \gls{ris} provides much wider dynamic ranges of $\sigma_1(\mathbf{H})$ and $\sigma_2(\mathbf{H})$ than diagonal \gls{ris}.
		Larger gaps are expected when the symmetry constraint can be relaxed.

@@

		\begin{subsubsection}{Rank-1 Indirect Channel}
			The indirect channel is rank-1 iff the forward or backward channel is rank-1.
			Let $\mathbf{H}^\mathrm{F} = \sigma^\mathrm{F} \mathbf{u}^\mathrm{F} {\mathbf{v}^\mathrm{F}}^\mathsf{H}$ without loss of generality.
			In this case, the channel Gram matrix can be written as Hermitian-plus-rank-1:
			\begin{equation}
				\mathbf{G} \triangleq \mathbf{H} \mathbf{H}^\mathsf{H} = \mathbf{Y} + \mathbf{z} \mathbf{z}^\mathsf{H},
			\end{equation}
			where $\mathbf{Y} \triangleq \mathbf{H}^\mathrm{D} (\mathbf{I} - \mathbf{v}^\mathrm{F} {\mathbf{v}^\mathrm{F}}^\mathsf{H}) \mathbf{H}^\mathrm{D} = \mathbf{T} \mathbf{T}^\mathsf{H}$ and $\mathbf{z} \triangleq \sigma^\mathrm{F} \mathbf{H}^\mathrm{B} \mathbf{\Theta} \mathbf{u}^\mathrm{F} + \mathbf{H}^\mathrm{D} \mathbf{v}^\mathrm{F}$.
			Regardless of \gls{ris} size and structure\footnote{A similar conclusion was made for diagonal \gls{ris} in \cite{Semmler2023}.}, its $n$-th ($n \ge 2$) eigenvalues are bounded by the Cauchy interlacing formula \cite{Golub2013}
			\begin{equation}
				\lambda_1(\mathbf{Y}) \ge {\color{blue}\lambda_2(\mathbf{G})} \ge \lambda_2(\mathbf{Y}) \ge \ldots \ge \lambda_{N-1}(\mathbf{Y}) \ge {\color{blue}\lambda_N(\mathbf{G})} \ge \lambda_N(\mathbf{Y}).
			\end{equation}
			The equivalent singular value inequality is
			\begin{equation}
				\sigma_1(\mathbf{T}) \ge {\color{blue}\sigma_2(\mathbf{H})} \ge \sigma_2(\mathbf{T}) \ge \ldots \ge \sigma_{N-1}(\mathbf{T}) \ge {\color{blue}\sigma_N(\mathbf{H})} \ge \sigma_N(\mathbf{T}).
				\label{iq:singular_value_interlacing}
			\end{equation}
			\eqref{iq:singular_value_interlacing} implies that, if the indirect channel is rank-1, then the \gls{ris} can at most enlarge the $n$-th ($n \ge 2$) channel singular value to the $(n-1)$-th singular value of $\mathbf{T}$.
			Note that the largest channel singular value is unbounded with a sufficiently large \gls{ris}.
		\end{subsubsection}

		\begin{subsubsection}{Fully-Connected \glsentryshort{ris} Without Direct Link}
			Denote the singular value decomposition of forward / backward channels as $\mathbf{H}^{\mathrm{B}/\mathrm{F}} = \mathbf{U}^{\mathrm{B}/\mathrm{F}} \mathbf{\Sigma}^{\mathrm{B}/\mathrm{F}} {\mathbf{V}^{\mathrm{B}/\mathrm{F}}}^\mathsf{H}$.
			The composite channel is
			\begin{equation}
				\mathbf{H} = \mathbf{H}^\mathrm{B} \mathbf{\Theta} \mathbf{H}^\mathrm{F} = \mathbf{U}^\mathrm{B} \mathbf{\Sigma}^\mathrm{B} \mathbf{X} \mathbf{\Sigma}^\mathrm{F} {\mathbf{V}^\mathrm{F}}^\mathsf{H},
			\end{equation}
			where $\mathbf{X} = {\mathbf{V}^\mathrm{B}}^\mathsf{H} \mathbf{\Theta} \mathbf{U}^\mathrm{F}$.
			% Since unitary matrices are closed under multiplication,
			\begin{proposition}
				In this case, the singular value bounds on $\mathbf{H}$ are equivalent to the singular value bounds on $\mathbf{BF}$, where $\mathbf{B}$ and $\mathbf{F}$ are arbitrary matrices with singular values $\mathbf{\Sigma}^\mathrm{B}$ and $\mathbf{\Sigma}^\mathrm{F}$.
				% The only singular value bounds applied here are those apply to the singular value bound of $\mathbf{F}\mathbf{B}$
			\end{proposition}
			\begin{proof}
				We first observe that singular value control problem can be solved w.r.t. unitary $\mathbf{X}$ and retrieved by $\mathbf{\Theta} = \mathbf{V}^\mathrm{B} \mathbf{X} {\mathbf{U}^\mathrm{F}}^\mathsf{H}$.
				Also, $\mathrm{sv}(\mathbf{U}^\mathrm{B} \mathbf{\Sigma}^\mathrm{B} \mathbf{X} \mathbf{\Sigma}^\mathrm{F} {\mathbf{V}^\mathrm{F}}^\mathsf{H}) = \mathrm{sv}(\bar{\mathbf{U}}^\mathrm{B} \mathbf{\Sigma}^\mathrm{B} \mathbf{\bar{V}}{{}^\mathrm{B}}^\mathsf{H} \bar{\mathbf{U}}^\mathrm{F} \mathbf{\Sigma}^\mathrm{F} \mathbf{\bar{V}}{{}^\mathrm{F}}^\mathsf{H}) = \mathrm{sv}(\mathbf{BF})$ where $\bar{\mathbf{U}}^{\mathrm{B}/\mathrm{F}}$ and $\bar{\mathbf{V}}^{\mathrm{B}/\mathrm{F}}$ are arbitrary unitary matrices.
			\end{proof}
			The problem now becomes, given $\mathbf{\Sigma}^\mathrm{B}$ and $\mathbf{\Sigma}^\mathrm{F}$, what can we say about the singular value of $\mathbf{BF}$.
			One comprehensive answer is Horn's inequality \cite{Bhatia2001}: for all admissible triples $(I, J, K)$,
			\begin{equation}
				\prod_{k \in {K}} \sigma_k(\mathbf{BF}) \leq \prod_{i \in {I}} \sigma_i(\mathbf{B}) \prod_{j \in {J}} \sigma_j(\mathbf{F}).
			\end{equation}
			It gives upper bound on the largest singular value and lower bound on the smallest singular value:
			\begin{align}
				\sigma_1(\mathbf{BF}) \le \sigma_1(\mathbf{B}) \sigma_1(\mathbf{F}) \label{iq:singular_value_largest}\\
				\sigma_N(\mathbf{BF}) \geq \sigma_N(\mathbf{B}) \sigma_N(\mathbf{F}) \label{iq:singular_value_smallest}.
			\end{align}
			Another useful result is introduced in \cite{Hogben2013}: for all $p>0$,
			\begin{equation}
				\sum_n \sigma_n^p(\mathbf{BF}) \leq \sum_n \sigma_n^p(\mathbf{B}) \sigma_n^p(\mathbf{F}) \label{iq:singular_value_power_sum}.
			\end{equation}
			When $p=2$, it implies the channel energy is upper bounded by the sum of element-wise power product of the forward and backward channels, as illustrated in Fig.~\subref*{sm:pc_power_bond_nd}.
			Interestingly, \eqref{iq:singular_value_largest}--\eqref{iq:singular_value_power_sum} are simultaneously tight when $\mathbf{X} = \mathbf{I}$ and $\mathbf{\Theta} = \mathbf{V}^\mathrm{B} {\mathbf{U}^\mathrm{F}}^\mathsf{H}$.
			This solution was claimed in \cite{Bartoli2023} to achieve channel capacity, but it is not true at moderate \gls{snr}.
			% In \cite{Bartoli2023}, the authors claimed this solution also achieves channel capacity but it is
		\end{subsubsection}
